R package qdist: A functional programming approach to truncated probability ditributions

### 2016-09-20 docs/qdist-article/truncdist.Rmd
qdist article revised

* probability distribution functions `p<dist>(q)` with `q` vector of quantiles;
* density functions `d<dist>(x)` with `x` vector of quantiles;
* quantile functions `q<dist>(p)` with `p` vector of probabilities;
* random number generation `r<dist>(n)` with `n` number of observations.

@@

We could easily write function `tpnorm()`, a truncated version for `pnorm()`, as:

@@

As an alternative approach, we decided to develop package `qdist` available at [https://github.com/andreaspano/qdist](https://github.com/andreaspano/qdist) .

@@

```{r, message=FALSE}
devtools::install_github("andreaspano/qdist", build=FALSE)
require(qdist)
tpnorm <- ptruncate("norm")
tdnorm <- dtruncate("norm")
tqnorm <- qtruncate("norm")
trnorm <- rtruncate("norm")
```

@@

The newly generated function `tpnorm()` has the same formals as the original `pnorm()`, plus two
extra parameters: `L` and `U`, respectively for lower and upper truncation thresholds set by default to
`-Inf` and `+Inf`:

```{r}
args(pnorm)
args(tpnorm)
```

Once we have defined `tpnorm()` we can use it for generating probability values from a non-
truncated normal distribution by leaving parameters `L` and `U` set to their defaults:

```{r}
q <- seq(6, 14, len = 100)
p_L_U <- tpnorm(q, mean = 10, sd = 1)
```

@@

```{r}
p_L9_U11 <- tpnorm(q, mean = 10, sd = 1, L = 9, U = 11)
```

@@

```{r, fig.width=8, warning=FALSE, message=FALSE}
df <- data.frame(
  q=q,
  tpnorm_type = factor(c(rep("p_L_U", times=length(q)), rep("p_L9_U11", times=length(q))), levels = c("p_L_U", "p_L9_U11")),
  value = c(p_L_U, p_L9_U11)
)

@@

pl <- ggplot(data = df, mapping = aes(x=q, y=value, col=tpnorm_type)) +
  geom_line(size=1) + scale_color_manual(values = c("p_L_U" = "darkgray", "p_L9_U11" ="darkred")) +   
  xlab("Quantile") + ylab("Probability") + 
  ggtitle("Probability plot for truncated and non-truncated normal distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=12), plot.title = element_text(size=17), axis.title=element_text(size=14))

@@

As a second example, we consider a density function for a left truncated Weibull distribution. We
first generate function `tdweibull()` by using `dtruncate()`:

```{r}
tdweibull <- dtruncate("weibull")
```

@@

```{r}
x <- qweibull(ppoints(1000), shape = 2, scale = 7)
d_L0 <- tdweibull(x, shape = 2, scale = 7)
d_L3 <- tdweibull(x, shape = 2, scale = 7, L= 3)
d_L5 <- tdweibull(x, shape = 2, scale = 7, L= 5)
d_L7 <- tdweibull(x, shape = 2, scale = 7, L= 7)
```

@@

```{r, warning=FALSE, message=FALSE}
df <- data.frame(
  x=x,
  tdweibull_type = factor(c(rep("d_L0", times=length(x)), rep("d_L3", times=length(x)), rep("d_L5", times=length(x)), rep("d_L7", times=length(x))), levels = c("d_L0", "d_L3", "d_L5", "d_L7")),
  value = c(d_L0, d_L3, d_L5, d_L7)
)

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=tdweibull_type)) +
  geom_line(size=1) + 
  scale_color_manual(values = c("d_L0" = "darkgray", "d_L3"="darkred", "d_L5" = "darkblue", "d_L7"="darkgreen")) +   
  xlab("Quantile") + ylab("Density") + ggtitle("Density plot for truncated Weibull distributions") +
  theme(legend.title=element_blank(), legend.text= element_text(size=12), plot.title = element_text(size=18), axis.title=element_text(size=14))
print(pl)
```

@@

```{r, warning=FALSE, message=FALSE}
require(evd)
trgumbel <- rtruncate("gumbel")
args(trgumbel)
rg <- trgumbel(10^4, loc = 3, scale = 2, L = 5)
df <- data.frame(rg=rg)

@@

```{r}
tqpois <- qtruncate("pois")
```

@@

```{r}
p <- ppoints(100)
# No truncation
q <- tqpois(p, lambda = 10)
# Lower truncation L = 9
q_L9 <- tqpois(p, lambda = 10, L = 9)
# Upper truncation H = 14
q_U14 <- tqpois(p, lambda = 10, U = 14)
# Both sides truncation L = 9, H = 14
q_L9_U14 <- tqpois(p, lambda = 10, L = 9, U = 14)
```

@@

```{r}
df <- data.frame(p=p,
  tqpois_type=factor(c(rep("No truncation", times=length(p)), rep("Lower truncation L = 9", times=length(p)), 
    rep("Upper truncation H = 14", times=length(p)), rep("Both sides truncation L = 9, H = 14", times=length(p))), 
    levels= c("No truncation", "Lower truncation L = 9", "Upper truncation H = 14", "Both sides truncation L = 9, H = 14")),
  value=c(q, q_L9, q_U14, q_L9_U14))

pl <- ggplot(data = df, mapping = aes(x=value, y=p, col=tqpois_type)) +
  geom_line(size=1, show.legend = FALSE) + 
  ylab("Probability") + xlab("Quantiles") + ggtitle("Quantiles from truncated Poisson distribution") +
  facet_wrap(~tqpois_type, nrow = 2) +
  theme(strip.text = element_text(size=12), plot.title = element_text(size=18), axis.title=element_text(size=14))

@@

```{r}
tqgamma <- qtruncate("gamma")
```

When we use `tqgamma()` with parameter rate it works but may return a set of `warnings`:

```{r}
tqgamma(.25, shape = 1, rate = .3)
```

@@

```{r, error=TRUE}
tqgamma(ppoints(10), shape = 1, scale = 3)
```

@@

```{r, error=TRUE}
tqgamma <- qtruncate("gamma")
tqgamma(p = .25, shape = 1, rate = 3)
```

@@

```{r}
tdnorm <- dtruncate("norm")
```

@@

```{r}
ltnorm <- function(x, L = -Inf, U = Inf) {
  theta <- c(mean(x), sd(x))
  ll <- function(theta, x, L = -Inf, U = Inf) {
    mean <- theta[1]
    sd <- theta[2]
    ld <- tdnorm(x = x, mean = mean, sd = sd, L = L , U = U, log = TRUE)
    -sum(ld)
  }
  optim(par = theta, fn = ll, x = x, L = L, U = U)[["par"]]
}
```

@@

```{r}
trnorm <- rtruncate("norm")
x <- trnorm(n = 1000, mean = 5, sd = 2, L = 3, U = 6)
ltnorm( x = x, L = 3, U = 6)
```

@@

>>>>

* probability distribution functions `p<dist>(q)` with `q` vector of quantiles;
* density functions `d<dist>(x)` with `x` vector of quantiles;
* quantile functions `q<dist>(p)` with `p` vector of probabilities;
* random number generator function `r<dist>(n)` with `n` number of observations.

@@

We could easily write function `ptnorm()`, a truncated version for `pnorm()`, as:

@@

As an alternative approach, we decided to develop package `qdist` available at [https://github.com/veronicagiro/qtruncate](https://github.com/veronicagiro/qtruncate) .

@@

```{r, message=FALSE}
devtools::install_github("veronicagiro/qtruncate", build=FALSE)
require(qdist)
ptnorm <- ptruncate("norm")
dtnorm <- dtruncate("norm")
qtnorm <- qtruncate("norm")
rtnorm <- rtruncate("norm")
```

@@

The newly generated function `ptnorm()` has the same formals as the original `pnorm()`, plus two
extra parameters: `L` and `U`, respectively for lower and upper truncation thresholds set by default to
`-Inf` and `+Inf`:

```{r}
args(pnorm)
args(ptnorm)
```

Once we have defined `ptnorm()` we can use it for generating probability values from a non-
truncated normal distribution by leaving parameters `L` and `U` set to their defaults:

```{r}
q <- seq(6, 14, len = 100)
p_L_U <- ptnorm(q, mean = 10, sd = 1)
```

@@

```{r}
p_L9_U11 <- ptnorm(q, mean = 10, sd = 1, L = 9, U = 11)
```

@@

```{r, fig.width=8, warning=FALSE, message=FALSE}
df <- data.frame(
  q=q,
  ptnorm_type = factor(c(rep("p_L_U", times=length(q)), rep("p_L9_U11", times=length(q))), levels = c("p_L_U", "p_L9_U11")),
  value = c(p_L_U, p_L9_U11)
)

@@

pl <- ggplot(data = df, mapping = aes(x=q, y=value, col=ptnorm_type)) +
  geom_line(size=1) + scale_color_manual(values = c("p_L_U" = "darkgray", "p_L9_U11" ="darkred")) +   
  xlab("Quantile") + ylab("Probability") + 
  ggtitle("Probability plot for truncated and non-truncated normal distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=12), plot.title = element_text(size=17), axis.title=element_text(size=14))

@@

As a second example, we consider a density function for a left truncated Weibull distribution. We
first generate function `dtweibull()` by using `dtruncate()`:

```{r}
dtweibull <- dtruncate("weibull")
```

@@

```{r}
x <- qweibull(ppoints(1000), shape = 2, scale = 7)
d_L0 <- dtweibull(x, shape = 2, scale = 7)
d_L3 <- dtweibull(x, shape = 2, scale = 7, L= 3)
d_L5 <- dtweibull(x, shape = 2, scale = 7, L= 5)
d_L7 <- dtweibull(x, shape = 2, scale = 7, L= 7)
```

@@

```{r, warning=FALSE, message=FALSE}
df <- data.frame(
  x=x,
  dtweibull_type = factor(c(rep("d_L0", times=length(x)), rep("d_L3", times=length(x)), rep("d_L5", times=length(x)), rep("d_L7", times=length(x))), levels = c("d_L0", "d_L3", "d_L5", "d_L7")),
  value = c(d_L0, d_L3, d_L5, d_L7)
)

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=dtweibull_type)) +
  geom_line(size=1) + 
  scale_color_manual(values = c("d_L0" = "darkgray", "d_L3"="darkred", "d_L5" = "darkblue", "d_L7"="darkgreen")) +   
  xlab("Quantile") + ylab("Density") + ggtitle("Density plot for truncated Weibull distributions") +
  theme(legend.title=element_blank(), legend.text= element_text(size=12), plot.title = element_text(size=18), axis.title=element_text(size=14))
print(pl)
```

@@

```{r, warning=FALSE, message=FALSE}
require(evd)
rtgumbel <- rtruncate("gumbel")
args(rtgumbel)
rg <- rtgumbel(10^4, loc = 3, scale = 2, L = 5)
df <- data.frame(rg=rg)

@@

```{r}
qtpois <- qtruncate("pois")
```

@@

```{r}
p <- ppoints(100)
# No truncation
q <- qtpois(p, lambda = 10)
# Lower truncation L = 9
q_L9 <- qtpois(p, lambda = 10, L = 9)
# Upper truncation H = 14
q_U14 <- qtpois(p, lambda = 10, U = 14)
# Both sides truncation L = 9, H = 14
q_L9_U14 <- qtpois(p, lambda = 10, L = 9, U = 14)
```

@@

```{r}
df <- data.frame(p=p,
  qtpois_type=factor(c(rep("No truncation", times=length(p)), rep("Lower truncation L = 9", times=length(p)), 
    rep("Upper truncation H = 14", times=length(p)), rep("Both sides truncation L = 9, H = 14", times=length(p))), 
    levels= c("No truncation", "Lower truncation L = 9", "Upper truncation H = 14", "Both sides truncation L = 9, H = 14")),
  value=c(q, q_L9, q_U14, q_L9_U14))

pl <- ggplot(data = df, mapping = aes(x=value, y=p, col=qtpois_type)) +
  geom_line(size=1, show.legend = FALSE) + 
  ylab("Probability") + xlab("Quantiles") + ggtitle("Quantiles from truncated Poisson distribution") +
  facet_wrap(~qtpois_type, nrow = 2) +
  theme(strip.text = element_text(size=12), plot.title = element_text(size=18), axis.title=element_text(size=14))

@@

```{r}
qtgamma <- qtruncate("gamma")
```

When we use `qtgamma()` with parameter rate it works but may return a set of `warnings`:

```{r}
qtgamma(.25, shape = 1, rate = .3)
```

@@

```{r, error=TRUE}
qtgamma(ppoints(10), shape = 1, scale = 3)
```

@@

```{r, error=TRUE}
qtgamma <- qtruncate("gamma")
qtgamma(p = .25, shape = 1, rate = 3)
```

@@

```{r}
dtnorm <- dtruncate("norm")
```

@@

```{r}
ltnorm <- function(x, L = -Inf, U = Inf) {
  theta <- c(mean(x), sd(x))
  ll <- function(theta, x, L = -Inf, U = Inf) {
    mean <- theta[1]
    sd <- theta[2]
    ld <- dtnorm(x = x, mean = mean, sd = sd, L = L , U = U, log = TRUE)
    -sum(ld)
  }
  optim(par = theta, fn = ll, x = x, L = L, U = U)[["par"]]
}
```

@@

```{r}
rtnorm <- rtruncate("norm")
x <- rtnorm(n = 10000, mean = 5, sd = 2, L = 3, U = 6)
ltnorm( x = x, L = 3, U = 6)
```

@@

### log, log.p and lower.tail formals

Unlike `truncdist` package, `qdist` allows us to use also the functionalities of formals `log` for density function, `log.p` and `lower.tail` for distribution and quantile functions.   
`qdist` functions uses `stats` distribution functions for computing from truncated distributions models, but not for performing the functionalities of `log`, `log.p` and `lower.tail` arguments, as they would have been incorrect applied to a truncated distribution. `log`, `log.p` and `lower.tail` functionalities are computed by using an alternative methods that produce results extremely close to the original ones.

<!--
Considering the Normal distribution as example, `ptruncate()` function uses `pnorm()` function to built probability function for truncated Normal distribution, defined as `ptnorm()`. However, `log.p` and `lower.tail` formals of the resulting function `ptnorm()` are not computed with the help of `pnorm()` function because their functionalities would have been incorrect applied to a truncated distribution. The functionalities of `log.p` and `lower.tail` are computed by using an alternative method that produces results extremely close to the original ones.

-->
